Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity

### 2022-01-24 manuscript_v7.Rmd
manuscript v7

## Key Points
- BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes from simulated SWOT data
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for unprecedented study of gas exchange spatiotemporal dynamics after SWOT's launch

@@

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020], i.e.e $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].

@@

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].

@@

@wangPhysicallyBasedScaling2021 recently attempted to address this problem by simulating *k* in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data.

@@

To answer this first question, we use the hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and then obtain a suite of models we empirically test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

@@

To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

@@

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. Scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

@@

### 2.3 Imposing large-river hydraulic geometry on *k* models
Given the theoretical context privded in Section 2.2, we know turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean * 100`% (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these flows as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models outlined above. While the complete model derivations for the small-eddy models (section 2.3.1) and the reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers. Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*

@@

![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Reynolds-extension model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Reynolds-extension model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg){width=1000px}

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

@@

For the $CO_2$ evasion and carbon efflux calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

@@

All of this algebra simplifies to equation 10. Conveniently, $k_{600}$ as measured by a tracer addition to a stream is inherently a reach-scale quantity. Therefore, equations 8 and 10 both yield a reach-scale $k_{600}$ (i.e. $k_{600_i}=k_{600}\forall i$). This lowers the number of parameters BIKER must infer and makes the problem much better constrained.

@@

With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

@@

With the setup described by equations 11 and 12, validation is performed using the BIKER posterior means. Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river. These metrics are defined in Table S2 and consist of the following:
- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
- Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

#### 3.2.3 Carbon efflux validation
It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere. Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the annual carbon emission rate from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. To do this, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 2, figure S1) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). Not all of the SWOT rivers have at least 14 observations and those rivers were removed for this experiment. Similarly, not all have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm. Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012. We then estimate a yearly average carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and then summing all of these emission rates to obtain an emissions rate in carbon per year across all rivers.

$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
Finally, we also compare BIKER's outputs with established in situ techniques, whose specific details and parameters are provided in Table S3. Streamflow is more readily modeled at the global scale than river channel geometry, and so these in situ models are used to predict $\bar{U}$ and *H* as functions of streamflow. Their outputs are in turn used to estimate *k* in upscaling workflows. Therefore, they rely on in situ streamflow records or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable). This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all three HG models). Of the three models tested here, 'Brinkerhoff2019' is trained on the dataset described in section 2.1, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].

### 3.3 Section 3 results
#### 3.3.1 BIKER
Figure 4 plots all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, again assuming no SWOT measurement error. Note that the y axis is normalized by maximum observed values to compare across rivers. In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER. Notably, there is sometimes positive bias in the estimates (e.g. Seine River). Some rivers are inferred nearly perfectly (e.g. Kanawha, Ohio Section 2, Wabash rivers) while others perform more poorly (e.g. Ash Slough, Ohio), though again much of this error stems from errors in inferring the magnitude of $k_{600}$ rather than its temporal dynamics. For some rivers, (e.g. Mississippi Intermediate), the first timestep is *THIS MIGHT BE DUE TO TO CERTAINTY OF THE PRIOR ETHIER ON K600 OR THE LIKELIHOOD???/*

@@

We calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5a, which plots the scores for rivers with and without SWOT measurement error (green and purple boxplots, respectively). For the no-error scenario, median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics. These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method. These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021]. Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, though the shape of the rBIAS boxplot in Figure 5a highlights a large portion of rivers which have a positive bias (Figure 4 also confirms this). Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.

![Figure 5. a: Performance metrics by river. See Table S2 for metric definitions. Dashed lines denote scores of 0, 0.50, and 1. CIs. b-c: Empirical cummulative density distributions of the *KGE* (b) and *r* (c) scores for both error and no-error rivers.](cache/validation/validation_by_river.jpg){width=1000px}

>>>>

## Key Points
- BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes from simulated SWOT data
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

@@

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].

@@

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though their focus was not on *k* and they relied on modeled streamflow to achieve this. This knowledge gap is contrasted against the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].

@@

@wangPhysicallyBasedScaling2021 recently attempted to address this problem by simulating *k* in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

@@

To answer this first question, we use hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a suite of models we empirically test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

@@

To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our dataset of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

@@

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

@@

### 2.3 Imposing large-river hydraulic geometry on *k* models
Given the theoretical context provided in Section 2.2, we know turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean * 100`% (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these flows as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models outlined above. While the complete model derivations for the small-eddy models (section 2.3.1) and the Reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers. Their derivations are outlined in sections 2.3.1 and 2.3.2 and their empirical tests are shown in section 2.4.*

@@

![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Reynolds-extension model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Reynolds-extension model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to further retard the dissipation rate via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores increase (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

@@

For the $CO_2$ evasion and carbon efflux calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Therefore, the raw carbon emissions estimates presented in this paper are largely meaningless in the context of actually measured carbon emissions from these rivers. This approach was deemed acceptable for this validation exercise because we are exclusively interested in the relative differences between methods tested and not the raw fluxes themselves.

@@

All of this algebra simplifies to equation 10. Conveniently, $k_{600}$ as measured by tracer additions to a stream is inherently a reach-scale quantity (in a mass-conserved reach). Therefore, equations 8 and 10 both yield a reach-scale $k_{600}$ (i.e. $k_{600_i}=k_{600}\forall i$). This lowers the number of parameters BIKER must infer and makes the problem much better constrained.

@@

With the sampling model described ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) specified (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

@@

With the setup described by equations 11 and 12, validation is performed for each river using the BIKER posterior means. These metrics are defined in Table S2 and consist of the following:

- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
- Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

#### 3.2.3 Carbon efflux validation
It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual carbon emitted from river to atmosphere. Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the annual carbon emission rate from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. To do this, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 2, figure S1) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). Not all of the SWOT rivers have at least 14 observations and those rivers were removed for this experiment. Similarly, not all have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm. Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012. We then estimate a yearly average carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and then summing all of these emission rates.
$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
Finally, we also compare BIKER's outputs with established in situ techniques, whose specific details and parameters are provided in Table S3. Streamflow is more readily modeled at the global scale than river channel geometry, and so these in situ models are used to predict $\bar{U}$ and *H* as functions of streamflow using hydraulic geometry (HG). Per equation 2, their outputs are in turn used to estimate *k*. Therefore, they rely on in situ streamflow records or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable). This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all three HG models). Of the three models tested here, 'Brinkerhoff2019' is trained on the dataset described in section 2.1, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].

### 3.3 Section 3 results
#### 3.3.1 BIKER
Figure 4 plots all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, again assuming no SWOT measurement error. Note that the y axis is normalized by maximum observed values to compare across rivers. In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER. Notably, there is sometimes positive bias in the estimates (e.g. Seine River). Some rivers are inferred nearly perfectly (e.g. Kanawha, Ohio Section 2, Wabash rivers) while others perform more poorly (e.g. Ash Slough, Ohio), though again much of this error stems from errors in inferring the magnitude of $k_{600}$ rather than its temporal dynamics. For some rivers, (e.g. Mississippi Intermediate), the first timestep is **THIS MIGHT BE DUE TO TO CERTAINTY OF THE PRIOR ETHIER ON K600 OR THE LIKELIHOOD?**

@@

We calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5a, which plots the scores for rivers with and without SWOT measurement error (green and purple boxplots, respectively). For the no-error scenario, median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics. These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is comparable to the better NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021]. Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a large portion of rivers which have a significant positive bias (Figure 4 also confirms this). Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.

![Figure 5. a) Performance metrics by river. See Table S2 for metric definitions. Dashed lines denote scores of 0, 0.50, and 1. b-c) Empirical cummulative density distributions of the *KGE* (b) and *r* (c) scores for both error and no-error rivers.](cache/validation/validation_by_river.jpg)

Figure 5 also compares BIKER results under a 'no-error' and 'error' scenario: the green and purple respectively in Figure 5a-c, where b-c plot the empirical cummulative density distributions (eCDFs) for *KGE* and *r* scores. There appears to be neglibile differences between results in Figures 5a-c, suggesting that SWOT measurement errors negligibly influence BIKER's performance across all four error metrics (Figure 5a). In fact, for some metrics, the error scenario appears to improve results. For example, median KGE without error is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE),2)` while with-error median KGE is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 1,]$KGE),2)`. However, caution should be used here as these purple boxplots are composed of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of the rivers in the green boxplots and many of the most poorly performing rivers are not in this set. The eCDFs more clearly indicate a marginal difference in performance when measurement errors are used to corrupt the BIKER inputs (with nearly identical performance in the best-performing rivers). Given these results, we deem that SWOT measurement error does not exert a significant influence on BIKER performance and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
